---
title: "*Mark*ing the fjord - Ecological assessment of small-scale grazing system along the fjord gradient in the Nordhordland UNESCO Biosphere Reserve"
format: html
# output-file: "drafts/NBRTradModThesis_IntroPreview.docx"
author:
  - name: Morgane Kerdoncuff
    orcid:
    email: morgane.kerdoncuf@uib.no
    affiliations:
      - Department of Biological Sciences, University of Bergen, Norway
      - Center for Sustainable Area Management (CeSAM), University of Bergen, Norway
date: today
bibliography: NBR_TRADMOD.bib
execute: 
  echo: false
---

# Scientific environment

This thesis was conducted in the Department of Biological Sciences, Faculty of Mathematics and Sciences, University of Bergen (Norway). I was also affiliated with the Centre for Sustainable Area Management (CeSAM), University of Bergen. This work was part of the TradMod project *From traditional resource use to modern industrial production: holistic management in Western Norway* funded by the Research Council of Norway (grant #280299). Additional funding support was provided by a grant from *Olaf Grolle Olsens's Legat til Universitet I Bergen med tilførsel av arv ette Miranda Bødker*.

# Acknowledgements

- Supervision: Inger, Amy
- Team colleagues: Jarrod, Alicia, Ieva, Katja
- Technician power quartermasters: Linn, Silje
- Stat & R jokers: Richard, Michael
- Office buddies: Mayke, Maaike, Joseph, Christian
- PhD buddies: Ragnhild, Camilla, Arnaud
- 1st season veterans: Micah, Astrid, Daiki
- 1st season veterans + bee workers: Ingrid, Kristi
- 2nd season veterans: Natalie, Christina, Anette
- 2nd season veteran + master evolution: Morgane
- 2nd season family time: Keith & Johan
- farmer jokers: Laila, 
- Nord family: Mie, Gabriel, Quentin

# Abstract in English

As scientific evidence demonstrates everyday the unprecedented environmental burden brought by agriculture worldwide, engaging research effort towards farming systems supporting agro-ecological diversity is an utmost priority. In Western Norway, agriculture is dominated by small-scale livestock production which has shaped semi-natural habitats of high ecological and cultural values across the whole fjord system. However the socio-economic decline of the countryside during past decades has led to concomitant trends of intensification on more fertile grasslands (infields) and abandonment of low-productive heathlands on the coast and in the mountains (outfields), increasing their vulnerability to environmental changes. In this context, the Nordhordland UNESCO Biosphere Reserve (NBR) aims to be a model area for sustainable land use that would stem from past management practices while ensuring the farmer livelihood and the resilience of the food production system.

In this thesis I relied on an exploratory ecological survey on the grazing fields of the NBR, across the whole coast-to-mountain gradient, to examine the potential of different managements for ensuring the delivery of multiple ecosystem services and enhancing local biodiversity. More specifically I attempted to assess both the influence of environmental variation driven by the fjord system geophysical context (PII, PIII, PIV) and the effects of land use including stocking density (PIII, PIV), livestock type (PII, PIII) and rotational prescribed burning (PI, PIV) on diversity, composition and associated ecosystem synergies and trade-offs in outfields (PI, PIV) and infields (PII, PIII, PIV).

I confirmed that the regional-scale environmental variation reflected a hump-shaped productivity gradient that affected the land use (PI, PIV) and the landscape (PIII). However when focusing on more productive infields I found no direct effect of landscape zone (PII, PIV) or climate variables (PIII) on grassland diversity (PII, PIV), assemblage composition (PIII) or ecosystem services (PII, PIV).

Regarding field management, I emphasized the importance of maintaining low-intensive grazing on outfields that showed very high bryophyte diversity and better carbon sequestration potential compared with infields (PIV). On the coast, rotational prescribed burning was a valuable complementary practice to year-round sheep grazing as it favoured xerophilous and sun-loving beetles by locally creating new ecological niches (PI). In infields, livestock type affected plant functional group, with on average higher percent cover of grass when grazed by cattle (PII) and higher percent cover of bryophyte when grazed by sheep (PIII). On the other hand stocking density or fertilisation level did not seem to affect the current diversity, composition and structure of grassland assemblages (PIII, PIV), but the intensification trend may have led to infield homogenisation over time (PIV).

*to continue* 

# Abstract in Norwegian

# List of publications

Publications included in the thesis:

- Paper I - **Kerdoncuff, M.**, Måren, I.E. & Eycott, A.E. (2023). Traditional prescribed burning of coastal heathland provides niches for xerophilous and sun-loving beetles. Biodiversity Conservation 32, 4083–4109. https://doi.org/10.1007/s10531-023-02684-x

- Paper II - Demeaux, M., **Kerdoncuff, M.**, Eycott, A. E., & Måren, I. E. (2024). Just graze it! Biodiversity, nectar and forage resources in cultural landscapes grazed by different livestock species. Ecosystems and People, 20(1). https://doi.org/10.1080/26395916.2024.2311176

- Paper III - **Kerdoncuff, M.**, Eycott, A.E, Klimes, A. & Måren, I.E. The farmers against the fjord: cross-scale interactions of environmental gradients and management on grassland assemblages in small-scale grazing systems. *Planned submission to Agriculture, Ecosystems and Environment*

- Paper IV - **Kerdoncuff, M.**, Eycott, A.E & Måren, I.E. Management of remote areas delivers complementary ecosystem services to more intensive grazing in semi-natural grasslands. *Planned submission to Journal of Applied Ecology*

# Author contributions

```{r AuthorContribution}
#| label: tbl-authorcontribution
#| tbl-cap: "Contributions to the individual papers in this thesis are categorised according to the CRediT (Contributor Roles Taxonomy) system."
#| message: false
#| warning: false

library(readxl)
library(gt)
library(tidyverse)

authorsummary <- read_xlsx("data/SummaryAuthContr.xlsx")
summary <- authorsummary |> 
  # group_by(Role) |> 
  pivot_wider(names_from = Paper, values_from = Contributor) |> 
  mutate_if(is.character, ~replace(., is.na(.), ""))

summary |>
  gt() |> 
  opt_stylize(style = 6, color = 'gray')
```

# Introduction - The role of Norwegian small-scale grazing systems in the current global food security challenge

Should we produce more to sustain more people on the planet? When facing the upcoming global food security challenge [@kastnerGlobalChangesDiets2012; @moraExploringFutureLand2020; @vanmeijlModellingAlternativeFutures2020], with a food demand expected to increase between 30% and 62% by 2050 [@vandijkMetaanalysisProjectedGlobal2021], the initial temptation to dive into a "crop-mania" has raised some warnings [@foleySolutionsCultivatedPlanet2011; @hunterAgriculture2050Recalibrating2017]. First, the prevalent issue is less a matter of gross production capacity than severe North-South inequalities in its redistribution and availability [@cassidyRedefiningAgriculturalYields2013; @fischerLandSparingLand2014; @tscharntkeGlobalFoodSecurity2012; @westLeveragePointsImproving2014]. Second, there is only very limited potential for the expansion of agricultural lands without conflicting with other land uses [@kimMappingEnvironmentalLand2018; @liSpatialConflictIdentification2023; @zabelGlobalAgriculturalLand2014] or damaging to a critical extent the biodiversity and functionality of natural and semi-natural ecosystems [@beckmannConventionalLandUse2019; @foleyGlobalConsequencesLand2005; @newboldGlobalEffectsLand2015; @newboldHasLandUse2016]. The livestock sector bears high responsibilities for the latter, as globalised grain-feed systems have steadily replaced pasture-based husbandry following the Green Revolution [@kastnerGlobalChangesDiets2012; @ramankuttyTrendsGlobalAgricultural2018]. This transition provided overall relatively poor return in calorie supply [@cassidyRedefiningAgriculturalYields2013] and economic profitability for the farmers [@garibaldiFarmingApproachesGreater2017; @jeanneretIncreaseFoodProduction2021], but was at the cost of massive land appropriation and severe environmental externalities in the Global South [@cunninghamCloseYieldgapSaving2013; @hongGlobalRegionalDrivers2021; @naylorLosingLinksLivestock2005] and of widespread decrease and degradation of global grassland area [@bardgettCombattingGlobalGrassland2021; @liuAnnualDynamicsGlobal2020; @piipponenGlobalTrendsGrassland2022; @wirseniusHowMuchLand2010], particularly sharp in Europe [@auneLossSeminaturalGrassland2018; @schilsPermanentGrasslandsEurope2022].

To shift away from this trajectory dramatically pressurising the Earth limits [@pelletierForecastingPotentialGlobal2010; @richardsonEarthSixNine2023], a new dominant paradigm has emerged during the past decade that acknowledges the synergy between food security, ecosystem integrity and people well-being and thus defends the potentiality of win-win alternatives [@fischerReframingFoodBiodiversity2017; @hanspachTradeoffsSynergiesFood2017; @wittmanSocialEcologicalPerspective2017]. These approaches have been declined across different concepts - e.g. agroecology [@holt-gimenezAgroecologyFoodSovereignty2012; @wezelAgroecologicalPrinciplesElements2020], diversified farming [@kremenDiversifiedFarmingSystems2012; @rosa-schleichEcologicaleconomicTradeoffsDiversified2019] or sustainable intensification [@prettySustainableIntensificationAgricultural2014; @rockstromSustainableIntensificationAgriculture2017] - that explore a wide range of management practices but share the consensus that food production should be diversified [@billenReshapingEuropeanAgrofood2021; @petersen-rockneyNarrowBrittleBroad2021] and tailored to the territorial context [@cunninghamCloseYieldgapSaving2013; @prettySustainableIntensificationAgricultural2014] to support self-sufficient and resilient socio-ecological systems [@chappellFoodSecurityBiodiversity2011; @gardnerAgriculturalResiliencePerspectives2019]. The realisation of this new agricultural pathway thus requires an extensive effort of trans-disciplinary and policy-oriented research experiments, that would explore ecological interactions and trade-offs under different farming systems [@egliWinnersLosersNational2018; @rockstromSustainableIntensificationAgriculture2017; @runhaarFourCriticalConditions2021].

In Western Norway, such opportunity has been offered through the UNESCO World Network of Biosphere Reserve [@unescoNewRoadmapMan2017], which supports the regional implementation of the Sustainable Development Goals by relying on wide-scale collaboration and knowledge-sharing between model areas disseminated all around the world [@barracloughGlobalKnowledgeAction2023]. As such, the Nordhordland UNESCO Biosphere Reserve (NBR) was designated to "be based on the best from the past and [...] pave the way for a future-oriented societal development that ensures the sustainable use of all types of resources for the benefit and pleasure of both the present inhabitants and future generations" [@kalandNordhordlandBiosphereReserve2018]. Agriculture in the NBR indeed carries a strong legacy of low-intensive rotational grazing that have shaped a typical mosaic-like landscape which hosted a remarkable biodiversity still unparalleled as of today [@austadHumaninfluencedVegetationTypes1991; @haugeGaldaneLaerdalWestern1988; @olssonAgroecosystemsNeolithicTime1991]. If the limiting landform and climate of the fjord system has prevented a too radical transformation of the countryside in comparison with the Eastern plains [@fjellstadPatternsChangeTwo1999], the rural communities still went through a lasting socio-economic decline leading to the intensification of fertile fields on one hand and the abandonment of remote, low-productive areas on the other [@forbordDriversChangeNorwegian2014; @hamreLandcoverStructuralChanges2007a]. As per the NBR objective, land planning for sustainable agriculture should therefore aim to 1) preserve the inherited ecological diversity associated with small-scale grazing and low-intensive management [@hamrePlantSpeciesRichness2010; @losvikHayMeadowCommunities1993] and 2) ensure the resilience and functionality of the agro-ecological system in face of upcoming global challenges such as climate change [@liuAssessingEffectsClimate2019; @meuwissenFrameworkAssessResilience2019]. However, despite its preponderance in the fjord landscape agriculture has so far received relatively low academic interest except when embedded in conservation research (but see for instance @wehnAdaptiveBiodiversityManagement2018), and extensive studies on grazed meadows particularly tend to be outdated (e.g. @losvikHayMeadowCommunities1993). The first stepping stone for coherent decision-making in the NBR management should thus be a comprehensive assessment of both the current ecological states of the NBR semi-natural habitats [@soliveresBiodiversityMultipleTrophic2016] and the direct and indirect benefits provided by the farmers activity [@bommarcoExploitingEcosystemServices2018; @goldsteinIntegratingEcosystemserviceTradeoffs2012].

<!-- The livestock sector holds key responsibilities as it accounts for the largest land use [@kastnerGlobalChangesDiets2012], global demand for animal products is still predicted to increase, even in the Global North [@alexandratosWorldAgriculture20302012], leading to further pressure over planetary boundaries [@newboldHasLandUse2016; @pelletierForecastingPotentialGlobal2010; @richardsonEarthSixNine2023]. Globalised grain-feed based systems have been growing relentlessly for the past 50 years following market demand, despite poor return in calorie and nutritional needs [@cassidyRedefiningAgriculturalYields2013; @ramankuttyTrendsGlobalAgricultural2018] and at the expense of severe environmental externalities mostly shouldered by the Global South [@cunninghamCloseYieldgapSaving2013; @hongGlobalRegionalDrivers2021; @jonesNutrientStrippingGlobal2013; @naylorLosingLinksLivestock2005]. In parallel pasture-based husbandry has either intensified or declined, leading to the decrease and degradation of global grassland area [@liNewGlobalLanduse2017; @liuAnnualDynamicsGlobal2020; @wirseniusHowMuchLand2010] and subsequent loss in productivity [@kwonGlobalEstimatesImpacts2016; @liuAssessingEffectsClimate2019; @piipponenGlobalTrendsGrassland2022], multifunctionality [@bardgettCombattingGlobalGrassland2021; @steffan-dewenterTradeoffsIncomeBiodiversity2007] and biodiversity [@newboldGlobalEffectsLand2015; @newboldHasLandUse2016]. -->
<!-- - such conversions sacrifice habitats of high conservation values [@bignalLowintensityFarmingSystems1996; @haladaWhichHabitatsEuropean2011]. As such, promoting pasture-based husbandry in the Global North would serve the complementary purposes of benefiting regional biodiversity while alleviating the crop burden of the Global South hotspots [@foleySolutionsCultivatedPlanet2011; @haberlQuantifyingMappingHuman2007; @kehoeGlobalPatternsAgricultural2015; @zabelGlobalImpactsFuture2019]. -->


::: {.callout-note icon=false}
## On labelling farming systems...

Characterising farming systems in an accurate and reproducible way is an everlasting challenge but nevertheless necessary to identify trade-offs brought by existing models in order to sort the wheat from the chaff. Since the environmental footprint of agricultural production became a burning topic, many different terms have been used usually to draw opposition between the intensive, industrialised model that became predominant in the Global North, usually branded as "conventional", and the more ecology-friendly farming qualified as "organic", "alternative" or "traditional" for the most famous [@sumbergWhatConventionalAgriculture2022]. However, most of them lack of meaningful conceptual definition: if organic farming is usually associated to regulated labels that allow a clear dichotomy based on well-defined parameters such as fertilisation or stocking density thresholds, "conventional" refers to what is common or generally in use in a certain place and at a certain time. Thus by essence "conventional" farming can vary a lot depending on countries and epochs, as for instance what can be currently described as "traditional" used to be "conventional" practices. Similarly, "traditional" farming is a very fuzzy concept that mostly refers to a socio-cultural context usually dated from before the Green Revolution and more or less arbitrarily associated with low-intensive management [@plieningerTraditionalLanduseNature2006]. 

In this thesis, "traditional" was used only when describing formerly common practices that have either severely declined or disappeared over time and are thus nowadays associated with the cultural heritage [@daugstadAgricultureUpholderCultural2006]. Instead I decided to characterise the NBR grazing system as "small-scale", which does not relate to specific production principles but to either spatial extent (up to 10 ha according to the @faoSmallholdersFamilyFarmers2013) or socio-economic status [@jouziOrganicFarmingSmallscale2017], and can therefore be more pertinent for comparative studies and wide-scale assessments [@ricciardiHowMuchWorlds2018]. I acknowledge that the terminology "small-scale" grazing is still lacking in the sense that it can encompass many different models and thus may not reflect as accurately the evolution of agriculture in the NBR. I therefore encourage further discussion as part of the NBR management plan to determine specific terminology for the past, current and predicted stages of farming in Western Norway that will facilitate policy-making in the future [@gillerFutureFarmingWho2021; @huberFarmTypologiesUnderstanding2024].
:::

# Aim of the thesis

This PhD thesis aimed to investigate how management practices and fjord environmental context affected the diversity, functionality and associated ecosystem services of the NBR grazing fields. As we wanted to provide a preliminary base of scientific knowledge that would guide and support decision-making in the NBR, we relied extensively on the ecosystem service framework that has grown familiar among the population and has been widely used for policy-oriented concepts (e.g. multifunctionality, see @bjorkhaugMultifunctionalAgriculturePolicy2008 and @wilsonWeakStrongMultifunctionality2008). Considering the limited scope of our ecological survey and the inherently high variability associated with the fjord geophysical context and the NBR small-scale grazing system, we took an explicit exploratory approach that relied on multiple indicators of ecological diversity and functioning (@tbl-papersummary).

Following a productivity gradient ranging from the coast to the mountain, we investigated:

- the effect of rotational prescribed burning on beetle diversity and ecological preferences by comparing active assemblages on pioneer and mature heath in a coastal heathland grazed all year round by Old Norse sheep (PI);
- the influence of livestock type (sheep, cow, goat) and landscape zone (coast, fjord) on the delivery of agronomic services in semi-natural grasslands by using indicators of nectar and forage resource based on plant functional traits (PII);
- the cross-scale interactions affecting the composition of these same semi-natural grasslands by testing correlations between sets of variables characterising fjord climate, landscape composition, grazing management, fine-scale environment and dominant grass, forb and beetle assemblages (PIII);
- the ecosystem services trade-offs and synergies provided by different land uses (infield, outfield) by assessing the aboveground and belowground productivity, the biodiversity and the carbon sequestration potential of grazing fields along the whole coast-to-mountain gradient (PIV).

```{r PaperSummary}
#| label: tbl-papersummary
#| tbl-cap: "Summary of explanatory (management and environment) and response (taxa and ecosystem attributes) variables directly -X- or indirectly -(X)- used in the papers of the thesis."
#| message: false
#| warning: false

library(readxl)
library(gt)
library(tidyverse)

papersummary <- read_xlsx("data/SummaryTablePapers.xlsx")
papersummary <- subset(papersummary, select = -c(Group))
summary <- papersummary |> 
  group_by(Subgroup) |> 
  pivot_wider(names_from = Paper, values_from = Included) |> 
  mutate_if(is.character, ~replace(., is.na(.), ""))

summary |>
  gt(groupname_col = "Subgroup") |> 
  cols_label(
    Variables = ""
    ) |> 
  opt_stylize(style = 6, color = 'gray')
```
